Spatial evolution of solar wind discontinuities in the outer heliosphere

### 2024-11-10 article.qmd
update: add discussion about Bn, PVI comparison, references to mission and clarification...

Discontinuities are not only an interesting topic in themselves but are also considered crucial for understanding the dynamics of the solar wind and cosmic rays. Theoretical models [@lerchePropagationMagneticDisturbances1975; @medvedevDissipativeDynamicsCollisionless1997] and MHD simulations [@grecoIntermittentMHDStructures2008; @yangFormationRotationalDiscontinuities2015] suggest that the formation and destruction of discontinuities are closely related to the nonlinear dynamics of Alfvén waves and Alfvénic turbulence. These nonlinear processes can create significant disturbances to the otherwise adiabatic evolution of the solar wind flow. 
These theoretical predictions are supported by spacecraft observations, which have revealed that the solar wind magnetic field follows the Parker model of the heliospheric current sheet only on average. Localized transient currents, which can be significantly more intense than those predicted by the Parker model, are carried by various discontinuities observed as strong variations in magnetic field components [@colburnDiscontinuitiesSolarWind1966; @burlagaMicroScaleStructuresInterplanetary1968; @turnerOrientationsRotationalTangential1971]. These discontinuities contribute significantly to the magnetic fluctuation spectra [@borovskyContributionStrongDiscontinuities2010] and play an important role in energy dissipation processes such as plasma heating [@osmanIntermittencyLocalHeating2012] and particle acceleration [@wentzelMotionMagneticDiscontinuities1964; @dmitrukTestParticleEnergization2004; @tesseinAssociationSuprathermalParticles2013]. 
More specifically, because tangential discontinuities act as boundaries separating different types of plasma and often involve oppositely directed magnetic fields on either side, they are potential regions where magnetic reconnection or merging can occur. Additionally, rotational discontinuities, being nonlinear large-amplitude Alfvén waves, are expected to significantly contribute to the diffusion coefficient for cosmic rays. Finally, these discontinuities can produce geomagnetic effects [@tsurutaniReviewInterplanetaryDiscontinuities2011] and impact space weather due to abrupt changes in external (solar wind) conditions.

In this study, we present a statistical analysis of discontinuities observed from Earth’s orbit to Jupiter, utilizing data collected by the Juno spacecraft [@boltonJunoMission2017] during its cruise phases (2011-2016), with its orbits shown in panel (a) of @fig-fig-overview. The primary goal of this paper is to investigate the radial dependence of the occurrence rate and other properties of discontinuities, in order to gain insights into their evolution and origin. Continuous measurements at 1 AU from missions such as Wind, ARTEMIS, and STEREO have been utilized to distinguish temporal effects from spatial variations by examining the characteristics of discontinuities at two radial distances simultaneously. This distinction is crucial because temporal variations could easily be misinterpreted as spatial variations [@tsurutaniInterplanetaryDiscontinuitiesTemporal1979]. Due to the potential misidentification of the normal direction of the magnetic field when using the minimum variance analysis (MVA) method [@hausmanDeterminingNatureOrientation2004; @liuFailuresMinimumVariance2023], we do not differentiate between the types of discontinuities, whether they are tangential, rotational, or either [@neugebauerReexaminationRotationalTangential1984].

@@

In this study we use datasets of four missions measuring solar wind magnetic field and plasma. Synergistic observations of these missions should advance our understanding of the solar wind discontinuities, their radial distribution and evolution [@velliUnderstandingOriginsHeliosphere2020].

<!-- ![Near-Earth's and STEREO-A solar wind plasma data during JUNO cruise phase](figures/omni_overview){#fig-overview} -->

Datasets collected from missions such as STEREO, ARTEMIS, and WIND will complement observations from Juno, offering a broader contextual perspective on heliospheric conditions and enabling a multi-point analysis of discontinuity properties [@velliUnderstandingOriginsHeliosphere2020]. The table below summarizes the time resolution of magnetic field and plasma data for Juno, Wind, ARTEMIS, and STEREO.

@@

We use Liu's [@liuMagneticDiscontinuitiesSolar2022] method to identify discontinuities in the solar wind. This method has better compatibility for the discontinuities with minor field changes, and is more robust to the situation encountered in the outer heliosphere. For each sampling instant $t$, we define three intervals: the pre-interval $[-1,-1/2]\cdot T+t$, the middle interval $[-1/,1/2]\cdot T+t$, and the post-interval $[1/2,1]\cdot T+t$, in which $T$ are time lags. Let time series of the magnetic field data in these three intervals are labeled ${\mathbf B}_-$, ${\mathbf B}_0$, ${\mathbf B}_+$, respectively. Then for an discontinuity, the following three conditions should be satisfied: (1) $\sigma({\mathbf B}_0) > 2\max\left(\sigma({\mathbf B}_-), \sigma({\mathbf B}_+)\right)$, (2) $\sigma\left({\mathbf B}_-+{\mathbf B}_+\right)>\sigma({\mathbf B}_-)+\sigma({\mathbf B}_+)$, and (3) $|\Delta {\mathbf B}|>|{\mathbf B}_{bg}|/10$, in which $\sigma$ and ${\mathbf B}_{bg}$ represent the standard deviation and the background magnetic field magnitude, and $\Delta {\mathbf B}={\mathbf B}(t+T/2)-{\mathbf B}(t-T/2)$. The first two conditions guarantee that the field changes of the discontinuity identified are large enough to be distinguished from the stochastic fluctuations or Alfven waves on magnetic fields, while the third is a supplementary condition to reduce the uncertainty of recognition.

Because the discontinuities may have different spatial scale at different radial distances from the Sun, we repeat the search procedure with time lags $T = 10, 20, 30, 40, 50, 60$ s, and combine the identified ID events together to establish the event list.
Discontinuities with spatial scales larger than $60$ s are not considered in this study, however it could be argued that they are not significant in the outer heliosphere if they follow the trend with decreasing number of IDs with increasing spatial scale. @fig-ljFit also indicates that the spatial scale of IDs does not vary significantly with the radial distance from the Sun.

TODO:
More than ? IDs are identified in the solar wind data from Juno, with ? IDs identified in 5 AU. Then for each discontinuity identified, we calculate the distance matrix of the time series sequence (distance between each pair of magnetic field vectors) to determine the leading edge and trailing edge of the discontinuity. After that, we use the minimum or maximum variance analysis (MVA) analysis [@sonnerupMinimumMaximumVariance1998; @sonnerupMagnetopauseStructureAttitude1967] to get the main (most varying) magnetic field component, $B_l$, and medium variation component, $B_m$. The maximum variance direction is then fitted by a hyperbolic tangent functions to extract the parameters to properly describe the discontinuity @eq-fit. Then we combine the magnetic field data and plasma data to obtain the thickness and the current density of the discontinuity.

@@

@@

![Examples of IDs from Juno, ARTEMIS, STEREO, and Wind](figures/fig_examples){#fig-examples}

@@

<!-- The basic approach of this proposal is to use solar wind measurements at 1AU (STEREO, ARTEMIS, WIND) to compare with Juno and PSP measurements and distinguish effects of solar wind temporal variations and effects of spatial (with the radial distance from the Sun) variations of discontinuity occurrence rate and characteristics. The example of such comparison for the occurrence rate is shown in @fig-rate, where we plot number of discontinuities measured per day by different spacecraft missions (for the same temporal resolution of magnetic field data and the same criteria of discontinuity determination). The radial distance of Juno for 2011-2015 is shown in @fig-model, and the number of discontinuities measured by Juno per day coincides with the discontinuity number measured by STEREO, WIND, and ARTEMIS, when Juno is around $1$ AU. This number (occurrence rate) decreases with distance (with time after $\sim 2013$), as Juno moves from $1$ AU to $5$ AU. We will use the similar comparison for discontinuity characteristics and occurrence rate derived for PSP and Juno.  -->

@@

The occurrence rate of IDs is investigated with regards of the radial distance from the Sun for Juno, ARTEMIS, STEREO, and Wind. Number of discontinuities measured per day by different spacecraft missions (for the same temporal resolution of magnetic field data and the same criteria of discontinuity determination) is shown in @fig-rate. Because data is not continuously available and ARTEMIS spacecrafts will only be exposed in the solar wind for a relative short time during its orbits around the Moon (the Moon would be in the magnetosphere for a long time as it orbits the earth), the occurrence rate (numbder of IDs per day) is corrected with respect to data gap and the time the mission spent in the pristine solar wind. 
The occurrence rate of IDs observed by Juno is consistent with the occurrence rate observed by 1AU missions, when Juno is around $1$ AU. This number (occurrence rate) decreases with distance as Juno moves from $1$ AU to $5$ AU (corresponding to time after $\sim 2013$, the radial distance of Juno for 2011-2015 is shown in @fig-model)

@@

![](figures/ocr/ocr_r_N2_cleaned)

@@

The peak current density $J_m$ of a discontinuity is determined by

$$
J_m = - \frac{1}{\mu_0 V_n} \frac{d B_m}{d t},
$$ {#eq-j}

where $\mu_0$ is the vacuum permeability and $V_n$ is the normal component of the proton flow velocity. For fitting method, the derivative of the magnetic field could be estimated as $d B_m / d t = B_{m0} / \Delta T$. This estimate is more robust than the direct calculation of the derivative of the magnetic field, because the fitting method is less sensitive to the noise in the data and the time resolution of the data (see discussion in the appendix).

@@

$$
d = (v \cdot n) \Delta T
$$

with $v = v_{sw} - v_{s/c} + U$ , where $\Delta T$ is the transition time (a fixed factor of the $\sigma$ used in fitting) , $v_{sw}$ is the solar wind velocity. The velocity of the spacecraft $v_{s/c}$ and the propagation speed of the discontinuities $U$ are negligible in our case.

To understand the relationship between the discontinuity and the local plasma, we normalize the thickness of the discontinuity by the initial ion length and the current density by the Alfven current density $J_A = e N_p V_A$ (corresponding to the drift between protons and electrons of local Alfén speed $V_A$). Figure @junoDistribution presents the statistical distribution of the (un)normalized thickness and current density of IDs grouped by the radial distance from the Sun collected by Juno. 
Panel (a,b) shows that the thickness increases with radial distance. However, after normalization, the thickness slightly decreases, with the most probable value around 4-2 $d_i$. With increasing radial distance, larger fraction of IDs have thickness less than 1 $d_i$, becombing sub-proton scale structures.  Panel (c,d) shows that the current density decreases with increasing radial distance. However, after normalization, it slightly increases, with the most probable value around 0.05-0.15 $J_A$. It should be noted that for the normalized thickness and current density, the change is not significant compared to the spread of the distribution.

@@

::: {#fig-ljFit}
![](figures/juno_distribution_r){#fig-ljFit}

![](figures/l_j_fit_xscale_log)

>>>>

Discontinuities are not only an interesting topic in themselves but are also considered crucial for understanding the dynamics of the solar wind and cosmic rays. Theoretical models [@lerchePropagationMagneticDisturbances1975; @medvedevDissipativeDynamicsCollisionless1997] and MHD simulations [@grecoIntermittentMHDStructures2008; @yangFormationRotationalDiscontinuities2015] suggest that the formation and destruction of discontinuities are closely related to the nonlinear dynamics of Alfvén waves and Alfvénic turbulence. These nonlinear processes can create significant disturbances to the otherwise adiabatic evolution of the solar wind flow. 
These theoretical predictions are supported by spacecraft observations, which have revealed that the solar wind magnetic field follows the Parker model of the heliospheric current sheet only on average. Localized transient currents, which can be significantly more intense than those predicted by the Parker model, are carried by various discontinuities observed as strong variations in magnetic field components [@colburnDiscontinuitiesSolarWind1966; @burlagaMicroscaleStructuresInterplanetary1968; @turnerOrientationsRotationalTangential1971]. These discontinuities contribute significantly to the magnetic fluctuation spectra [@borovskyContributionStrongDiscontinuities2010] and play an important role in energy dissipation processes such as plasma heating [@osmanIntermittencyLocalHeating2012] and particle acceleration [@wentzelMotionMagneticDiscontinuities1964; @dmitrukTestParticleEnergization2004; @tesseinAssociationSuprathermalParticles2013]. 
More specifically, because tangential discontinuities act as boundaries separating different types of plasma and often involve oppositely directed magnetic fields on either side, they are potential regions where magnetic reconnection or merging can occur. Additionally, rotational discontinuities, being nonlinear large-amplitude Alfvén waves, are expected to significantly contribute to the diffusion coefficient for cosmic rays. Finally, these discontinuities can produce geomagnetic effects [@tsurutaniReviewInterplanetaryDiscontinuities2011] and impact space weather due to abrupt changes in external (solar wind) conditions.

In this study, we present a statistical analysis of discontinuities observed from Earth’s orbit to Jupiter, utilizing data collected by the Juno spacecraft [@boltonJunoMission2017] during its cruise phases (2011-2016), with its orbits shown in panel (a) of @fig-overview. The primary goal of this paper is to investigate the radial dependence of the occurrence rate and other properties of discontinuities, in order to gain insights into their evolution and origin. Continuous measurements at 1 AU from missions such as Wind, ARTEMIS, and STEREO have been utilized to distinguish temporal effects from spatial variations by examining the characteristics of discontinuities at two radial distances simultaneously. This distinction is crucial because temporal variations could easily be misinterpreted as spatial variations [@tsurutaniInterplanetaryDiscontinuitiesTemporal1979]. Due to the potential misidentification of the normal direction of the magnetic field when using the minimum variance analysis (MVA) method [@hausmanDeterminingNatureOrientation2004; @liuFailuresMinimumVariance2023], we do not differentiate between the types of discontinuities, whether they are tangential, rotational, or either [@neugebauerReexaminationRotationalTangential1984].

@@

In this study, we analyze Juno observations during its cruise phase from August 25, 2011, to July 30, 2016, spanning approximately 1,800 days. Data collected from missions such as STEREO, ARTEMIS, and WIND complement Juno’s observations, providing a broader perspective on heliospheric conditions and enabling a multi-point analysis of discontinuity properties [@velliUnderstandingOriginsHeliosphere2020]. Specifically, we utilize magnetic fields from Juno [@boltonJunoMission2017], measured by the Juno Fluxgate Magnetometer (MAG) [@connerneyJunoMagneticField2017], alongside ion bulk velocity and plasma density with hourly resolution from a solar wind model at Juno’s location (described below). For ARTEMIS [@angelopoulosARTEMISMission2011], we use magnetic field data from the Fluxgate Magnetometer [@austerTHEMISFluxgateMagnetometer2008] and ion bulk velocity and plasma density derived from the Electrostatic Analyzer [@mcfaddenTHEMISESAPlasma2009]. For STEREO[@kaiserSTEREOMissionIntroduction2008], we incorporate magnetic fields from the magnetic field experiment on IMPACT [@acuñaSTEREOIMPACTMagnetic2008; @luhmannSTEREOIMPACTInvestigation2008] and ion bulk velocity and plasma density from PLASTIC [@galvinPlasmaSuprathermalIon2008]. For WIND, we include magnetic fields from the Magnetic Field Investigation (MFI) [@leppingWINDMagneticField1995] and ion bulk velocity and plasma density from the Solar Wind Experiment (SWE) [@ogilvieSWEComprehensivePlasma1995]. The table below summarizes the time resolution of magnetic field and plasma data for each mission. These synergistic observations are expected to enhance our understanding of solar wind discontinuities, including their radial distribution and evolution.

@@

We employ Liu’s method [@liuMagneticDiscontinuitiesSolar2022] to identify discontinuities in the solar wind. This approach provides better compatibility for discontinuities with small magnetic field changes and demonstrates greater robustness for conditions in the outer heliosphere compared to traditional methods, namely the B-criterion [@burlagaDirectionalDiscontinuitiesInterplanetary1969] and the TS-criterion [@tsurutaniInterplanetaryDiscontinuitiesTemporal1979]. For each sampling time $t$, we define three intervals: the pre-interval $[-1,-1/2]\cdot T+t$, the middle interval $[-1/2,1/2]\cdot T+t$, and the post-interval $[1/2,1]\cdot T+t$, where $T$ represents time lags. The magnetic field timeseries in these three intervals are labeled as ${\mathbf B}_-$, ${\mathbf B}_0$, and ${\mathbf B}_+$, respectively. To identify a discontinuity, the following three conditions must be met:

$$
\begin{aligned}
\sigma({\mathbf B}_0) &> 2\max\left(\sigma({\mathbf B}_-), \sigma({\mathbf B}_+)\right) \\
\sigma\left({\mathbf B}_-+{\mathbf B}_+\right) &>\sigma({\mathbf B}_-)+\sigma({\mathbf B}_+) \\
|\Delta {\mathbf B}| &>|{\mathbf B}_{bg}|/10
\end{aligned}
$$

Here, $\sigma$ denotes the standard deviation, ${\mathbf B}_{bg}$ represents the background magnetic field magnitude, and $\Delta {\mathbf B}={\mathbf B}(t+T/2)-{\mathbf B}(t-T/2)$. The first two conditions ensure that the field changes are distinguishable from stochastic fluctuations, while the third condition serves as a supplementary measure to reduce recognition uncertainty.

Notably, this method shares certain similarities with the Partial Variance of Increments (PVI) approach for identifying coherent structures, as both calculate indices based on standard deviations (with this method focusing on the standard deviation of the magnetic field, and PVI on magnetic field increments). However, this method is better suited for discontinuity identification by comparing data within neighboring time windows rather than across a correlation scale (an unknown quantity that may vary with radial distance and solar wind conditions). As a result, this approach effectively excludes wave-like coherent structures, eliminating the need for manual inspection and bypassing the necessity of grouping continuous clusters of high-PVI points as in PVI analysis.

Because the discontinuities may have different spatial scale at different radial distances from the Sun, we repeat the search procedure with time lags $T = 10, 20, 30, 40, 50, 60$ s, and combine the identified ID events together to establish the event list.
Discontinuities with spatial scales larger than $60$ s are not considered in this study, however it could be argued that they are not significant in the outer heliosphere if they follow the trend with decreasing number of IDs with increasing spatial scale. @fig-junoDistribution also indicates that the spatial scale of IDs does not vary significantly with the radial distance from the Sun.

<!-- TODO: -->

More than ? IDs are identified in the solar wind data from Juno, with ? IDs identified in 5 AU. Then for each discontinuity identified, we calculate the distance matrix of the time series sequence (distance between each pair of magnetic field vectors) to determine the leading edge and trailing edge of the discontinuity. After that, we use the minimum or maximum variance analysis (MVA) analysis [@sonnerupMinimumMaximumVariance1998; @sonnerupMagnetopauseStructureAttitude1967] to transform the magnetic field into the boundary normal (LMN) coordinate system. The main (most varying) magnetic field component, $B_l$, is then fitted by a hyperbolic tangent functions to extract the parameters to properly describe the discontinuity @eq-fit. Then we combine the magnetic field data and plasma data to obtain the thickness and the current density of the discontinuity.

@@

Previous studies have indicated that the minimum variance direction obtained through Minimum Variance Analysis (MVA) is frequently unreliable. Improved accuracy can be achieved by either using the cross product of the magnetic field measured at the boundaries as the normal direction [@knetterFourpointDiscontinuityObservations2004] or by enforcing conditions such as $Δ|B|/|B| > 0.05$ or $ω > 60°$ [@liuFailuresMinimumVariance2023]. The former approach requires a small normal component in the magnetic field, while the latter approach limits the dataset to a subset that satisfies these conditions. In our analysis, we applied both methods and observed no significant differences in the results.

<!-- TODO: use both methods to determine the normal direction of the discontinuity and compare the results. -->

@@

::: {#fig-examples}
![Examples of IDs from Juno, ARTEMIS, STEREO, and Wind](figures/fig_examples)

Examples of solar wind discontinuities observed by various spacecraft at different heliocentric distances and times. Panels (a)–(c) display data from Juno at 1 AU (a), 3 AU (b), and 5 AU (c), showing magnetic field components $B_l$ (blue), $B_n$ (red), $B_m$ (green), and the magnetic field magnitude $B_{\text{total}}$ (black). Panels (d)–(f) show magnetic field and ion speed observations from other spacecraft: ARTEMIS on March 27, 2012 (d), STEREO on June 28, 2016 (e), and Wind on August 26, 2011 (f).
:::

@@

<!-- The basic approach of this proposal is to use solar wind measurements at 1AU (STEREO, ARTEMIS, WIND) to compare with Juno and PSP measurements and distinguish effects of solar wind temporal variations and effects of spatial (with the radial distance from the Sun) variations of discontinuity occurrence rate and characteristics. The example of such comparison for the occurrence rate is shown in @fig-rate, where we plot number of discontinuities measured per day by different spacecraft missions (for the same temporal resolution of magnetic field data and the same criteria of discontinuity determination). The radial distance of Juno for 2011-2016 is shown in @fig-model, and the number of discontinuities measured by Juno per day coincides with the discontinuity number measured by STEREO, WIND, and ARTEMIS, when Juno is around $1$ AU. This number (occurrence rate) decreases with distance (with time after $\sim 2013$), as Juno moves from $1$ AU to $5$ AU. We will use the similar comparison for discontinuity characteristics and occurrence rate derived for PSP and Juno.  -->

@@

The occurrence rate of IDs is investigated with regards of the radial distance from the Sun for Juno, ARTEMIS, STEREO, and Wind. Number of discontinuities measured per day by different spacecraft missions (for the same temporal resolution of magnetic field data and the same criteria of discontinuity determination) is shown in @fig-rate. Because data is not continuously available and ARTEMIS spacecrafts will only be exposed in the solar wind for a relative short time during its orbits around the Moon (the Moon would be in the magnetosphere for a long time as it orbits the earth), the occurrence rate (numbder of IDs per day) is corrected with respect to data gap and the time the mission spent in the pristine solar wind. 
The occurrence rate of IDs observed by Juno is consistent with the occurrence rate observed by 1AU missions, when Juno is around $1$ AU. This number (occurrence rate) decreases with distance as Juno moves from $1$ AU to $5$ AU (corresponding to time after $\sim 2013$, the radial distance of Juno for 2011-2016 is shown in @fig-model)

@@

![](figures/ocr/ocr_r_N2_cleaned_simplified.pdf)

@@

The current density $J_m$ across a discontinuity is determined by

$$
J_m = - \frac{1}{μ_0 V_n} \frac{d B_l}{d t},
$$ {#eq-j}

where $μ_0$ is the vacuum permeability and $V_n$ is the normal component of the proton flow velocity. For fitting method, the derivative of the magnetic field could be estimated as $\max(d B_l / d t) = B_{l0} / \Delta T$. This estimate is more robust than the direct calculation of the derivative of the magnetic field, because the fitting method is less sensitive to the noise in the data and the time resolution of the data (see discussion in the appendix).

<!--  The proton flow velocity $v$ is expressed as $v = v_{sw} - v_{s/c} + U$, where $v_{sw}$ is the solar wind velocity, $v_{s/c}$ is the spacecraft velocity, and $U$  denotes the propagation speed of the discontinuities. In our case, the latter two velocities are negligible. -->

@@

$$
d = V_n \Delta T
$$

where $\Delta T$ is the transition time derived from the fitting method.

To understand the relationship between the discontinuity and the local plasma, we normalize the thickness of the discontinuity by the initial ion length and the current density by the Alfven current density $J_A = e N_p V_A$ (corresponding to the drift between protons and electrons of local Alfén speed $V_A$). Figure @fig-junoDistribution presents the statistical distribution of the (un)normalized thickness and current density of IDs grouped by the radial distance from the Sun collected by Juno. 
Panel (a,b) shows that the thickness increases with radial distance. However, after normalization, the thickness slightly decreases, with the most probable value around 4-2 $d_i$. With increasing radial distance, larger fraction of IDs have thickness less than 1 $d_i$, becombing sub-proton scale structures.  Panel (c,d) shows that the current density decreases with increasing radial distance. However, after normalization, it slightly increases, with the most probable value around 0.05-0.15 $J_A$. It should be noted that for the normalized thickness and current density, the change is not significant compared to the spread of the distribution.

@@

::: {#fig-junoDistribution}
![](figures/juno_distribution_r)
